---
title: "Capital Bikeshare — Data Dashboard"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    self_contained: true
    toc: true
    toc_float:
      collapsed: false
    fig_width: 12
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  cache.path = "cache/"
)
options(tigris_use_cache = TRUE)
dir.create("cache", showWarnings = FALSE)
source("R/analysis.R")
```

---

## Data availability

The heatmap below shows every year-month period that has been successfully
downloaded and processed.  Colour intensity reflects trip volume; grey cells
indicate periods where no file was processed.

```{r availability-chart, fig.height=7}
avail <- summarize_data_availability()

if (nrow(avail) == 0L) {
  cat("> **No processed data found.**",
      "Run `run_pipeline(all_available = TRUE)` first.")
} else {
  knitr::kable(
    avail |>
      dplyr::summarise(
        `First period` = min(paste0(year, "-", sprintf("%02d", month))),
        `Last period`  = max(paste0(year, "-", sprintf("%02d", month))),
        `Periods`      = dplyr::n(),
        `Total trips`  = scales::comma(sum(n_trips, na.rm = TRUE))
      ),
    caption = "Coverage summary"
  ) |> print()

  print(build_availability_chart(avail))
}
```

---

## Trip density by census tract

Trip **start** coordinates are spatially joined to 2020 US Census tracts
covering the Capital Bikeshare service area (Washington D.C., Maryland, and
Virginia).  Tract boundaries are downloaded via the
[`tigris`](https://github.com/walkerke/tigris) package using the data's
own geographic extent.  Colour is log-scaled.

```{r tract-map, fig.width=10, fig.height=10, cache=TRUE, eval=Sys.getenv("GITHUB_ACTIONS") != "true"}
ext <- tryCatch(
  trip_extent(),
  error = function(e) {
    message("Could not compute extent: ", conditionMessage(e),
            " — using default service-area bounding box.")
    list(xmin = -77.6, xmax = -76.8, ymin = 38.7, ymax = 39.2)
  }
)

tracts_sf <- tryCatch(
  download_census_tracts(extent = ext, year = 2020L),
  error = function(e) {
    message("Census tract download failed: ", conditionMessage(e))
    NULL
  }
)

if (is.null(tracts_sf)) {
  cat("> **Census tract data unavailable** — tigris download failed.",
      "Check your internet connection and retry.")
} else {
  tracts_counts <- tryCatch(
    aggregate_trips_to_tracts(tracts_sf = tracts_sf, sample = 1000000L),
    error = function(e) {
      message("Trip aggregation failed: ", conditionMessage(e))
      tracts_sf$n_trips <- NA_integer_
      tracts_sf
    }
  )

  print(build_tract_map(tracts_counts))
}
```

---

## Per-year trip counts

```{r yearly-table}
if (nrow(avail) > 0L) {
  yearly <- avail |>
    dplyr::group_by(year) |>
    dplyr::summarise(
      months_available = dplyr::n(),
      total_trips      = sum(n_trips, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::mutate(total_trips = scales::comma(total_trips))

  knitr::kable(
    yearly,
    col.names = c("Year", "Periods available", "Total trips"),
    caption   = "Annual trip totals (processed periods only)"
  )
}
```

---

---

> **Interactive version:** a fully interactive dashboard with per-year
> choropleth maps, monthly bar charts, and live filtering is available at
> [**interactive.html**](interactive.html) (generated by
> `analysis_interactive.py` using
> [pygris](https://github.com/walker-data/pygris) census tract boundaries).

*Generated automatically by the
[Capital Bikeshare pipeline](https://github.com/Bryvado/CapitalBikeshareData).*
